\chapter{Implementation}
\label{chap:implementation}

This chapter documents the concrete implementation of the architecture
described in Chapter~\ref{chap:architecture}. Section~\ref{sec:stack} lists
the technology choices and justifies them briefly. Sections~\ref{sec:backend-layering}
through~\ref{sec:gemini-integration-impl} focus on the backend:
layered organisation, PostgreSQL persistence, ChromaDB integration, and
Gemini API wrapping. Section~\ref{sec:frontend-impl} covers the React
frontend, and Section~\ref{sec:deployment-impl} describes the Docker Compose
deployment.

\section{Technology Stack}
\label{sec:stack}

The implementation is written in Python 3.12 on the backend and TypeScript
on the frontend. The main libraries and their roles are summarised in
Table~\ref{tab:stack}.

\begin{table}[h]
\centering
\caption{Main technology choices and their role in the system.}
\label{tab:stack}
\begin{tabular}{l l}
\hline
\textbf{Component} & \textbf{Library / version} \\
\hline
Web framework         & FastAPI 0.115 \\
ASGI server           & uvicorn 0.30 \\
ORM                   & SQLAlchemy 2.0 \\
Configuration         & pydantic-settings 2.5 \\
Password hashing      & passlib + bcrypt \\
JWT                   & python-jose 3.3 \\
PDF parsing           & PyMuPDF 1.24 \\
Vector database       & ChromaDB 0.5 (persistent) \\
LLM and embeddings    & google-generativeai 0.8 (Gemini) \\
Relational database   & PostgreSQL 16 \\
Testing               & pytest 8.3, pytest-cov, pytest-mock \\
Frontend build        & Vite 8, TypeScript 6 \\
UI framework          & React 19 \\
Styling               & Tailwind CSS 4 \\
UI components         & shadcn/ui (Radix-based) \\
HTTP client           & axios 1.15 \\
Routing               & React Router 7 \\
Container orchestration & Docker Compose (v5 plugin) \\
\hline
\end{tabular}
\end{table}

The choice of FastAPI is driven by three factors: built-in asynchronous
support, automatic OpenAPI generation, and first-class integration with
pydantic for request and response validation. The choice of ChromaDB over
alternatives such as Qdrant, Weaviate, or Pinecone is driven by its
embeddable mode: ChromaDB runs in-process with a persistent client and a
simple directory on disk, which keeps the system self-contained and easy to
reproduce. Gemini was chosen because it offers a generous free tier that
covers both embeddings (\texttt{text-embedding-004}) and generation
(\texttt{gemini-1.5-flash}), which is sufficient for both development and
the experimental evaluation reported in Chapter~\ref{chap:evaluation}.

\section{Backend Layering}
\label{sec:backend-layering}

The backend follows a strict layered organisation inspired by the
repository and service patterns common in Spring and NestJS:

\begin{verbatim}
backend/
  api/
    routes/      # FastAPI routers; HTTP-level only
    dependencies.py  # DI wiring, JWT middleware
  services/      # business logic
  repositories/  # data access (PostgreSQL + ChromaDB)
  models/
    schemas.py     # Pydantic request/response shapes
    db_models.py   # SQLAlchemy ORM classes
  core/
    config.py      # pydantic-settings configuration
    constants.py   # chunk size, top-k, model identifiers, ...
    database.py    # engine, sessionmaker, get_db dependency
    logging.py     # root logger setup
    security.py    # bcrypt hashing, JWT encode/decode
  main.py          # FastAPI app, lifespan hook, router wiring
  tests/
    unit/
    integration/
    conftest.py    # SQLite in-memory override, TestClient fixture
\end{verbatim}

Routes are intentionally thin: they validate the incoming payload through
a pydantic schema, call the appropriate service, and translate
domain-specific exceptions into HTTP status codes. Services contain the
business logic and depend only on abstract repositories, which in turn are
the only components that know about SQLAlchemy sessions or ChromaDB
clients. This separation keeps the services trivially testable with
in-memory SQLite or mocked ChromaDB clients.

Dependency injection is implemented with FastAPI's \texttt{Depends}
mechanism. For example, the authentication service is wired as follows:

\begin{verbatim}
def get_user_repository(session: Session = Depends(get_db)):
    return UserRepository(session)

def get_auth_service(repo: UserRepository = Depends(get_user_repository)):
    return AuthService(repo)

@router.post("/login")
def login(payload: UserLoginRequest,
          auth_service: AuthService = Depends(get_auth_service)):
    ...
\end{verbatim}

In tests, the \texttt{get\_db} dependency is overridden with a function
that yields a session bound to an in-memory SQLite engine (using
SQLAlchemy's \texttt{StaticPool} so that every connection shares the same
database), and the rest of the DI graph is assembled automatically by
FastAPI.

\section{PostgreSQL Persistence}
\label{sec:postgres-impl}

The relational schema is a single \texttt{users} table, mapped by the
\texttt{User} ORM class. Its columns are defined with the SQLAlchemy 2.0
typed \texttt{Mapped}/\texttt{mapped\_column} API, which provides
compile-time type information and seamless integration with IDEs and
type-checkers.

Tables are created on application startup with a single call to
\texttt{Base.metadata.create\_all}, executed inside the FastAPI
\texttt{lifespan} hook. This is simpler than a full migration framework
such as Alembic and is sufficient for a single-table schema that does not
evolve. The same startup hook calls
\texttt{AuthService.seed\_demo\_users}, which inserts the three demo
accounts (\texttt{avocatura@demo.com}, \texttt{restaurant@demo.com},
\texttt{admin@demo.com}) if they are not already present. The method is
idempotent, so restarting the container is a safe no-op after the first
run.

Password hashing uses bcrypt through passlib with a cost factor of 12,
which corresponds to approximately 250\,ms of CPU time per hash on a
modern laptop. This cost is high enough to make brute-force attacks
expensive but low enough to keep login latency below the threshold of
perception.

\section{JWT Authentication}
\label{sec:jwt-impl}

The JWT layer is implemented in \texttt{core/security.py} and exposed
through the \texttt{get\_current\_user} FastAPI dependency.
Tokens are signed with HS256 and a 32-byte secret loaded from the
environment; the claim set is kept minimal --- a subject
(\texttt{sub=<user\_id>}), an issued-at timestamp, an expiration
timestamp, and a copy of the user's email for logging purposes. On every
protected request, the dependency decodes the token, verifies its
signature and expiration, looks up the user in PostgreSQL, and injects
the resulting \texttt{User} row into the handler. A missing, malformed,
or expired token results in a 401 response with the appropriate
\texttt{WWW-Authenticate} header, preserving HTTP semantics.

The authentication module is test-driven. Every behaviour --- successful
registration, duplicate email, valid login, wrong password, unknown
email, missing token, invalid token, protected route --- is exercised by
a dedicated unit or integration test. The current test suite contains 22
tests and reaches 83\,\% branch coverage across the backend, with
\texttt{auth\_service} at 96\,\%, \texttt{user\_repository} at 100\,\%,
and \texttt{api/routes/auth} at 100\,\%.

\section{Document Ingestion with PyMuPDF}
\label{sec:ingestion-impl}

PDF handling is delegated to PyMuPDF (imported as \texttt{fitz}), which
is substantially faster and more reliable than pure-Python alternatives
such as PyPDF2. The ingestion service opens the uploaded file with
\texttt{fitz.open(stream=...)}, iterates over the pages, and calls
\texttt{page.get\_text("text")} to obtain the textual content of each
page in reading order.

The resulting text is split into chunks by a pure-Python helper. The
chunking strategy is a sliding window of approximately 500 tokens with
50 tokens of overlap between consecutive chunks. Token counts are
approximated by splitting on whitespace --- this is conservative (actual
subword-tokenised counts are usually slightly higher) but avoids
importing a full tokenizer in the backend, which would add hundreds of
megabytes to the container image.

Each chunk produced by the chunker is accompanied by a metadata
dictionary containing the original filename, the page number it was
extracted from, and a shared \texttt{document\_id}, which is a UUID4
generated once per document. These dictionaries flow unchanged through
the rest of the pipeline, all the way to the final citation in the
response.

\section{Vector Store Integration}
\label{sec:vector-store-impl}

ChromaDB is used in persistent mode with a single directory on disk as
the store. The directory is mounted as a Docker volume
(\texttt{/data/chroma}) so that restarts preserve the indexed data. The
\texttt{VectorStore} repository encapsulates every call to the ChromaDB
API and exposes three methods to the service layer:

\begin{itemize}
  \item \texttt{add\_chunks(user\_id, chunks, embeddings, metadatas)},
        which upserts a batch of chunks into the user's collection;
  \item \texttt{query(user\_id, query\_embedding, top\_k)}, which
        returns a list of \texttt{(chunk\_text, SourceCitation)} tuples;
  \item \texttt{delete\_document(user\_id, document\_id)}, which
        removes every chunk sharing the given \texttt{document\_id}
        from the user's collection.
\end{itemize}

Collections are created lazily on first access. Their names follow the
pattern \texttt{user\_<id>}, with \texttt{<id>} taken from the user's
primary key in PostgreSQL; this guarantees that the namespace of
collections is exactly the namespace of user ids and prevents any
accidental cross-user access.

\section{Gemini API Integration}
\label{sec:gemini-integration-impl}

The Gemini integration is encapsulated in \texttt{EmbeddingService}
(used during ingestion and for query embedding) and inside
\texttt{QueryService} (used for answer generation). Both wrap the
\texttt{google.generativeai} SDK behind a narrow interface, which makes
them trivial to mock in unit tests and simplifies the eventual
replacement of Gemini with another LLM provider.

Two Gemini models are used:

\begin{itemize}
  \item \texttt{models/text-embedding-004} --- a 768-dimensional dense
        embedding model optimised for retrieval. It is applied to both
        document chunks during ingestion and to queries at search time.
  \item \texttt{models/gemini-1.5-flash} --- a fast, large-context
        generative model with a context window of up to 1\,million
        tokens. For the present system the prompt rarely exceeds a few
        thousand tokens, but the large window leaves room for future
        experiments with higher values of $k$.
\end{itemize}

Both the embedding and the generation calls are guarded by explicit
error handling: API failures are caught, logged with their payload, and
surfaced as HTTP 502 (``bad gateway'') to the client, so that the
frontend can show a dedicated error message instead of an opaque 500.

\section{Frontend Implementation}
\label{sec:frontend-impl}

The frontend is a Vite-based React 19 application written in TypeScript.
Styling is done with Tailwind CSS~4 using CSS variables for theme
colours; shadcn/ui provides the component primitives (button, card,
dialog, badge, sidebar, skeleton, etc.) as source files that are copied
into the project rather than imported from a package, giving full
ownership of the component code.

The top-level \texttt{App} component defines a React Router tree with
the following routes:

\begin{itemize}
  \item \texttt{/login} and \texttt{/register} --- unauthenticated
        entry points, each a single centred card;
  \item \texttt{/dashboard} --- the authenticated home, listing the
        user's uploaded documents and exposing an upload dialog;
  \item \texttt{/chat} --- the RAG interface, with a chat log, a
        question input, and source badges rendered below every answer;
  \item \texttt{/how-to-use} --- a Romanian-language help page aimed
        at end users, explaining how to upload a PDF, how to ask
        questions, how to read the answer and its citations, and what
        the current limitations are.
\end{itemize}

A \texttt{ThemeProvider} at the root of the tree manages dark mode by
toggling a class on the \texttt{html} element; the preference is
persisted in \texttt{localStorage} under a dedicated key. HTTP
communication with the backend goes through a single axios instance
that reads the base URL from a Vite environment variable and
automatically attaches the stored JWT to every outgoing request.

\section{Deployment with Docker Compose}
\label{sec:deployment-impl}

Three services are defined in \texttt{docker-compose.yml}:
\texttt{postgres}, \texttt{backend}, and \texttt{frontend}. PostgreSQL
runs unchanged from the official \texttt{postgres:16-alpine} image and
is exposed on host port 5433, as discussed in
Section~\ref{sec:deployment-model}. The backend uses a Python 3.12-slim
base image, installs the Python dependencies from
\texttt{requirements.txt}, copies the application source, and runs
uvicorn on port 8000. The frontend is built in a first Docker stage
with Node 20, then its \texttt{dist/} folder is copied into an
\texttt{nginx:alpine} image and served on port 80; an nginx
\texttt{location /api/} block proxies requests to the backend service
inside the Docker network.

Environment variables are provided through a single \texttt{.env} file
at the root of the \texttt{rag\_application/} folder, which is ignored
by git. The \texttt{.env.example} file committed to the repository
lists every expected variable with a placeholder value, so that a new
developer can clone the repository, copy the example, fill in the
Gemini API key, and run \texttt{docker compose up --build} to get a
fully working stack.
