\chapter{System Architecture}
\label{chap:architecture}

This chapter describes the architecture of the proposed RAG system at a
level of abstraction that is independent of the specific technologies used
to implement it. Concrete technology choices --- FastAPI, SQLAlchemy,
ChromaDB, Google Gemini, React, Docker --- are discussed in the next chapter.

\section{Overview}
\label{sec:architecture-overview}

The system follows a classical three-tier structure. A single-page web
application running in the user's browser issues HTTP requests to a backend
REST API. The backend is organised in layers that separate transport,
business logic, and persistence; it owns two distinct storage back-ends: a
relational database for user accounts and session data, and a vector
database for document chunk embeddings. A third external dependency --- the
Google Gemini API --- provides both the embedding model used during
ingestion and the generative model used at query time.

Five actors participate in every user interaction:

\begin{enumerate}
  \item the \textbf{browser}, which serves the React application and holds
        the JWT access token in \texttt{localStorage};
  \item the \textbf{backend}, which receives HTTP requests, validates
        tokens, and orchestrates the business logic;
  \item the \textbf{relational database} (PostgreSQL), which stores user
        records;
  \item the \textbf{vector database} (ChromaDB), which stores embeddings
        together with their metadata;
  \item the \textbf{external LLM service} (Gemini), which computes
        embeddings and generates the final answers.
\end{enumerate}

The backend is the only component allowed to talk to the databases and the
LLM service. The browser never holds the API key of Gemini, never connects
to ChromaDB directly, and never sees another user's data.

\section{Ingestion Pipeline}
\label{sec:ingestion}

The ingestion pipeline turns a raw PDF upload into a set of indexed
embeddings. Its five stages are:

\begin{enumerate}
  \item \textbf{Upload.} The authenticated user submits a PDF through an
        HTTP POST request that carries the file as a multipart form and the
        JWT access token in the \texttt{Authorization} header. The backend
        validates the content type (\texttt{application/pdf}) and rejects
        files larger than a configurable maximum (20\,MB by default).
  \item \textbf{Text extraction.} PyMuPDF opens the PDF in memory and
        produces, for every page, a string containing its textual content.
        Pages that are empty or contain only scanned images without OCR are
        skipped with a warning.
  \item \textbf{Chunking.} Each page's text is split into overlapping
        chunks of approximately 500 tokens, with an overlap of 50 tokens
        between consecutive chunks. The overlap ensures that a semantic
        unit straddling the chunk boundary is never lost at retrieval
        time. Each chunk carries the page number in its metadata.
  \item \textbf{Embedding.} The chunks of a document are sent to the Gemini
        embedding model in batches, producing one dense vector per chunk.
        Batching reduces both latency and the number of API calls, which
        matters in the free-tier quota used for development.
  \item \textbf{Indexing.} The chunks, their metadata, and their embedding
        vectors are appended to the ChromaDB collection dedicated to the
        uploading user. The collection name follows a deterministic pattern
        (\texttt{user\_<id>}) that makes it trivial for the backend to
        route subsequent queries to the right collection.
\end{enumerate}

Once the pipeline completes, the backend returns a \texttt{DocumentMetadata}
response containing the internal document id, the original filename, the
number of pages, and the number of resulting chunks. The metadata is stored
in ChromaDB as part of each chunk, so that listing and deletion can be
implemented purely through vector store operations, without a separate
PostgreSQL table for documents.

\section{Query Pipeline}
\label{sec:query-pipeline}

The query pipeline implements the RAG recipe described in
Section~\ref{sec:rag}. Given a natural-language question from a logged-in
user, it executes the following steps:

\begin{enumerate}
  \item \textbf{Authentication.} The \texttt{Authorization: Bearer <token>}
        header is validated by the JWT middleware. If the token is missing,
        expired, or tampered with, the request is rejected with HTTP
        401 before any other work is done.
  \item \textbf{Query embedding.} The user's question is passed to the
        same Gemini embedding model used during ingestion, producing a
        single query vector.
  \item \textbf{Similarity search.} ChromaDB is queried for the top-$k$
        nearest chunks inside the current user's collection. With
        $k=5$, experimental experience shows that the retrieved set almost
        always contains the evidence needed to answer the question, while
        remaining small enough to fit in the LLM's prompt budget.
  \item \textbf{Prompt construction.} The backend assembles a prompt
        consisting of a system instruction (\texttt{answer using only the
        context below; if the answer is not in the context, say so}),
        the retrieved chunks in order, and the original question. Each
        retrieved chunk is prefixed with a short header containing its
        filename and page number so that the model can refer to them.
  \item \textbf{Generation.} The prompt is sent to the Gemini generative
        model, which returns a natural-language answer.
  \item \textbf{Citation extraction.} The backend inspects the metadata of
        the retrieved chunks and produces a list of
        \texttt{(filename, page)} pairs, which is returned to the client
        alongside the answer text.
\end{enumerate}

The final JSON response has the shape

\begin{verbatim}
{
  "answer": "The probation period is 90 days per the
             collective contract.",
  "sources": [
    {"filename": "contract_colectiv_2024.pdf", "page": 12},
    {"filename": "regulament_intern.pdf",      "page":  4}
  ]
}
\end{verbatim}

\noindent which is the format consumed by the frontend chat interface.

\section{Data Model}
\label{sec:data-model}

The relational part of the data model is intentionally minimal. A single
table, \texttt{users}, holds the authenticated accounts. Its columns are
the auto-incremented primary key \texttt{id}, the unique \texttt{email},
the \texttt{hashed\_password} computed with bcrypt, and the
\texttt{created\_at} timestamp. No document metadata is stored in
PostgreSQL; documents live entirely in the vector database. This decision
simplifies the implementation but would need to be revisited in a
production system that requires features such as document sharing or
server-side search over filenames.

The vector store uses one ChromaDB collection per user. Every chunk stored
in a collection carries three mandatory metadata fields:

\begin{itemize}
  \item \texttt{document\_id} --- a UUID assigned when the document is
        ingested; identical for every chunk of the same document.
  \item \texttt{filename} --- the original name of the uploaded file.
  \item \texttt{page} --- the page number the chunk was extracted from
        (1-indexed, as is conventional in PDF tooling).
\end{itemize}

Because each collection is named after a specific user id, no additional
filter is needed at query time: membership in a collection is itself the
authorisation check.

\section{Authentication and User Isolation}
\label{sec:auth-architecture}

Authentication is handled by a JWT middleware built on top of FastAPI's
\texttt{OAuth2PasswordBearer}. The flow is standard:

\begin{enumerate}
  \item The user registers with an email and a password. The password is
        hashed with bcrypt (cost factor 12) before being stored in
        PostgreSQL.
  \item The user logs in with the same credentials. The backend verifies
        the password against the stored hash and, on success, issues a
        signed JWT with a 24-hour lifetime. The token's \texttt{sub}
        claim contains the user id, and its payload also includes the
        email for logging purposes.
  \item On every subsequent request to a protected endpoint, the
        middleware decodes and validates the JWT, loads the user row
        from PostgreSQL, and injects it into the request handler as a
        dependency.
\end{enumerate}

Data isolation rests on two complementary mechanisms. First, at the HTTP
layer, every route that accesses documents or issues queries depends on
\texttt{get\_current\_user}: a request without a valid token is rejected
before any business logic is executed. Second, at the vector store
layer, collections are named after user ids, so a query executed on
behalf of user \texttt{A} cannot read chunks belonging to user \texttt{B}
even if a programming error at the HTTP layer were to occur. This is a
form of defence in depth: the architecture makes the safe case the
only case that can actually run.

\section{Deployment Model}
\label{sec:deployment-model}

The entire stack is packaged as a Docker Compose project with three
services: PostgreSQL, the FastAPI backend, and an nginx container serving
the compiled React frontend. Volumes are used to persist the PostgreSQL
data directory, the ChromaDB store, and the uploaded PDFs. The PostgreSQL
container is exposed on a non-standard host port (5433 instead of 5432)
so that it does not conflict with a native PostgreSQL instance that may
already be running on the developer's machine. Details of the
containerisation are given in Section~\ref{sec:deployment-impl}.
